---
author: Saul, Jhet, Frankie, and Chase
title: "Testicular Cancer EDA"
format:
  pdf:
    warning: false
editor: 
  markdown: 
    wrap: 72
---

# Research Questions:

-   Github Repo: <https://github.com/Smoorad99/485-TT>

The questions we aim to solve with our data are the following:

-   How does race affect survival rate/months survived for testi cancer?

    -   Exploring the role race plays in testicular cancer survival
        rates could help us address disparities between different races.
        If we find that survival rate of testicular cancer is impacted
        by race, it would push us to explore why this may be. For
        example, it may indicate inequality in healthcare received by
        different races.

-   How does survival rate/months change based on access to treatment?

    -   Exploring the relationship between survival rate/months and time
        from diagnosis to treatment helps us understand the effects of
        delaying treatment after diagnosis. Investigating the quality of
        life patients experience while undergoing different treatments
        may also help us better understand the effectiveness of each
        treatment.

-   How does the survival rate/months change based on the marital status
    of the patient?
Abstract:

Testicular cancer is one of the most common types of cancer in both adolescent and young adult males. Fortunately, it has a high five-year survival rate of about 95%. The data used in our analysis comes from the Surveillance, Epidemiology, and End Results Program (SEER) database. Multiple survival analysis methodologies were applied including Kaplan-Meier estimator and Cox proportional hazards regression. Key factors such as race, marital status, months from diagnosis to treatment, and order in which radiation and surgery were performed were all included in the analysis. Our model was found to be significant by three tests. It was found treatment, diagnosis timing, marital status, and race to be significant in predicting survival outcomes among those with testicular cancer. Survival analysis serves as a significant tool in investigating the history of testicular cancer. The findings of this study can assist in clinical decision making, improving patient care, and ultimately working towards a greater survival rate for those with this disease.

# Data Management

```{r}
library(dplyr)
library(ggplot2)
library(survminer)
library(survival)
```

```{r}
#df <- read.csv("C:/Users/cjhet/Documents/testis.csv") 
df <- read.csv("C:/data/testi.csv")
# This data is a subset of our original data only inclusinf observations whose primary site was on their testicles.

df[df == "Blank(s)"] <- NA
```


```{r}
# Jhet's DM
df <- df %>% 
    mutate(
    Sex = as.factor(Sex),
    Scope.Reg.LN.Sur = as.factor(RX.Summ..Scope.Reg.LN.Sur..2003..),
    Surg.Oth.Reg.Dis = as.factor(RX.Summ..Surg.Oth.Reg.Dis..2003..),
    Surg.Rad.Seq = as.factor(ifelse(RX.Summ..Surg.Rad.Seq == "No radiation and/or cancer-directed surgery", "None", "Surg and or rad")),
    Systemic.Sur.Seq = as.factor(RX.Summ..Systemic.Sur.Seq..2007..),
    Months.diag.to.treat = ifelse(as.numeric(Months.from.diagnosis.to.treatment) == 0, "<1", "1+")
) %>% select(-c(3:12))
```

```{r}
#Chase's DM
df <- df %>% 
  mutate(tumor_size = as.numeric("CS Tumor Size/Ext Eval (2004-2015)"))
```

```{r}
# Frakie's DM
df <- select(df, -c('Number.of.Examined.Pelvic.Nodes.Recode..2010..':'Lymph.vascular.Invasion..2004..varying.by.schema.'))

names(df)[names(df) == "CS.tumor.size..2004.2015."] <- "CS_tumor_size"
names(df)[names(df) == "Regional.nodes.examined..1988.."] <- "Regional_node_examined"
names(df)[names(df) == "Regional.nodes.positive..1988.."] <- "Regional_positive_node"
names(df)[names(df) == "CS.lymph.nodes..2004.2015."] <- "CS_lymph_node"
names(df)[names(df) == "CS.mets.at.dx..2004.2015."] <- "CS_mets_dx"
```

```{r}
df$CS_tumor_size <- as.numeric(as.character(df$CS_tumor_size))

# Create the new variable 'tumor_size_mod' based on the conditions
df$tumor_size_mod <- ifelse(df$CS_tumor_size == 000, 0, # No mass/tumor
                                ifelse(df$CS_tumor_size >= 001 & df$CS_tumor_size <= 988, df$CS_tumor_size, # Exact size in mm
                                ifelse(df$CS_tumor_size == 989, 989, # 989 mm or larger
                                ifelse(df$CS_tumor_size == 990, 0.5, # Microscopic, no given size, assume 0.5mm as a placeholder
                                ifelse(df$CS_tumor_size == 991, 5, # Less than 1 cm, use 5mm as the midpoint
                                ifelse(df$CS_tumor_size == 992, 15, # Less than 2 cm, use 15mm as the midpoint
                                ifelse(df$CS_tumor_size == 993, 25, # Less than 3 cm, use 25mm as the midpoint
                                ifelse(df$CS_tumor_size == 994, 35, # Less than 4 cm, use 35mm as the midpoint
                                ifelse(df$CS_tumor_size == 995, 45, # Less than 5 cm, use 45mm as the midpoint
                                ifelse(df$CS_tumor_size >= 996 & df$CS_tumor_size <= 998, NA, # Site-specific codes, handle as NA or a specific rule if applicable
                                ifelse(df$CS_tumor_size == 999 | df$CS_tumor_size == 888 | df$CS_tumor_size == 1022, NA, # Unknown, Not applicable, or Blank, handle as NA
                                NA))))))))))) # Default case to catch any unhandled values
```

```{r}
#modifying regional nodes examined 

# Step 1: Create a new numeric variable for the exact number of nodes examined
df$nodes_examined_num <- ifelse(df$Regional_node_examined >= 01 & df$Regional_node_examined <= 89, df$Regional_node_examined,
                                    ifelse(df$Regional_node_examined == 90, 90, NA))

# Step 2: Create a new categorical variable for special conditions
df$nodes_examined_cat <- ifelse(df$Regional_node_examined == 00, "No nodes examined",
                                    ifelse(df$Regional_node_examined == 95, "Aspiration performed",
                                    ifelse(df$Regional_node_examined == 96, "Sampling, number unknown",
                                    ifelse(df$Regional_node_examined == 97, "Dissection, number unknown",
                                    ifelse(df$Regional_node_examined == 98, "Removed, number unknown",
                                    ifelse(df$Regional_node_examined == 99, "Unknown",
                                    ifelse(df$Regional_node_examined == 126, "Blank", "Exact number")))))))

```

```{r}

#modifying regional nodes positive

# Step 1: Numeric variable for the exact number of positive nodes
df$positive_nodes_num <- ifelse(df$Regional_positive_node >= 1 & df$Regional_positive_node <= 89, df$Regional_positive_node,
                                    ifelse(df$Regional_positive_node == 90, 90, NA))

# Step 2: Categorical variable for special conditions
df$positive_nodes_cat <- ifelse(df$Regional_positive_node == 0, "All nodes negative",
                                    ifelse(df$Regional_positive_node == 95, "Positive aspiration performed",
                                    ifelse(df$Regional_positive_node == 97, "Positive, number unspecified",
                                    ifelse(df$Regional_positive_node == 98, "No nodes examined",
                                    ifelse(df$Regional_positive_node == 99, "Unknown",
                                    ifelse(df$Regional_positive_node == 126, "Blank", NA)))))) # Using NA for 'Exact number positive' to keep only special conditions

# Optional: Convert the categorical variable into a factor for easier handling in certain types of analyses
df$positive_nodes_cat <- factor(df$positive_nodes_cat)
```

```{r}
#Saul's DM
df$`Age.recode.with.single.ages.and.90.` <- gsub(
  " years","",df$`Age.recode.with.single.ages.and.90.`)
df <- df %>% 
  mutate(
    age = as.numeric(`Age.recode.with.single.ages.and.90.`),
    race = `Race.recode..W..B..AI..API.`,
    survival.months = as.numeric(Survival.months),
    marital.status.at.diagnosis = Marital.status.at.diagnosis,
    situ.malignant.tumors = as.numeric(`Total.number.of.in.situ.malignant.tumors.for.patient`),
    tm.3.1998.2003 = `Tumor.marker.3..1998.2003.`,
    id = Patient.ID,
    death.site = COD.to.site.recode.ICD.O.3.2023.Revision.Expanded..1999..,
    cause.specific = SEER.cause.specific.death.classification,
    cause.other = SEER.other.cause.of.death.classification)

# Other recodes
df <- df %>% mutate(
  situ.malignant.tumors = ifelse(situ.malignant.tumors == 1, "One tumor",
                                 ifelse(situ.malignant.tumors > 1,
                                        "More than one tumor", NA)),
  tm.3.1998.2003 = recode(tm.3.1998.2003, 
                          "0" = "None Done (SX)",
                          "2" = "Within normal limits (S0)",
                          "4" = "Range 1 (S1) <1.5 x upper limit of normal for LDH assay",
                          "5" = "Range 2 (S2) 1.5 – 10 x upper limit of normal for LDH assay",
                          "6" = "Range 3 (S3) >10 x upper limit of normal for LDH assay",
                          "8" = "Ordered, results not in chart",
                          "9" = "Unknown or no information",
                          "14" = " Blank")
)

df <- df %>% select(id, age,
                    race, death.site, cause.specific,
                    cause.other, survival.months,
                    marital.status.at.diagnosis, 
                    situ.malignant.tumors,
                    tm.3.1998.2003,
                    Sex,
                    Scope.Reg.LN.Sur,
                    Surg.Oth.Reg.Dis,
                    Surg.Rad.Seq,
                    Systemic.Sur.Seq,
                    Months.diag.to.treat,CS_tumor_size,
                    Regional_node_examined,
                    Regional_positive_node,
                    CS_lymph_node,
                    CS_mets_dx,
                    tumor_size_mod,
                    nodes_examined_num,
                    Regional_node_examined,
                    nodes_examined_cat)
```

# Exploratory Data Analysis

```{r}
summary(df)
```

## Jhet

After cleaning and managing the data, all of the variables I am in
charge of analyzing are now categorical, meaning statistics such as
mean, standard deviation and median are not very useful.

Instead, we can see the frequency distributions above in count form, as
well as visualized below.

```{r}
#50522 rows
table(df$Scope.Reg.LN.Sur)
ggplot(df, aes(x=Scope.Reg.LN.Sur)) +
  geom_bar(aes(fill=Scope.Reg.LN.Sur)) +
  theme(axis.text.x=element_blank(),
        legend.text=element_text(size=6))
```

This variable contains data relating to the scope of surgery done to
regional lymph nodes. Most of the cases are "none", meaning no surgery
was done. Besides NA, the next highest number of observations is in the
4+ lymph nodes removed category, followed by 1-3.

From this we can discern that removal of lymph nodes is uncommon, but
becomes more necessary the more positive nodes there are.

<br/>

```{r}
table(df$Surg.Oth.Reg.Dis)
ggplot(df, aes(x=Surg.Oth.Reg.Dis)) +
  geom_bar(aes(fill=Surg.Oth.Reg.Dis)) +
  theme(axis.text.x=element_blank(),
        legend.text=element_text(size=6))
```

Surg.Oth.Reg.Dis stands for "surgery to other distant regions", and over
\~42,000 out of 50,000 observations are in the "none/diagnosed at
autopsy" category.

Another 6000 of the remaining 8000 are NA's, meaning the variable
contains little meaningful data for our investigations.

<br/>

```{r}
table(df$Surg.Rad.Seq)
ggplot(df, aes(x=Surg.Rad.Seq)) +
  geom_bar(aes(fill=Surg.Rad.Seq)) +
  theme(axis.text.x=element_blank(),
        legend.text=element_text(size=6))
```

The categories in this variable originally contained the order of
surgery and or radiation. However, \~40,000 were again contained in the
"no surgery" category.

Thus, the variable is collapsed into just two levels, one in which no
surgery or radiation was used, and another where one or both were used.

<br/>

```{r}
table(df$Systemic.Sur.Seq)
ggplot(df, aes(x=Systemic.Sur.Seq)) +
  geom_bar(aes(fill=Systemic.Sur.Seq)) +
  theme(axis.text.x=element_blank(),
        legend.text=element_text(size=6))
```

Similar to the previous variable, this one contains data involving the
sequence of systemic surgery and therapy. Most of the observations are
in the none level, another large portion are patients that got systemic
therapy after surgery, and another large chunk is NA's.

The presence of two substantial categories makes this a useful variable
for our treatment related data science question.

<br/>

```{r}
table(df$Months.diag.to.treat)
ggplot(df, aes(x=Months.diag.to.treat)) +
  geom_bar(aes(fill=Months.diag.to.treat))
```

Just like the Surg.Rad.Seq variable, almost all of the patients received
treatment within 1 month of their diagnosis.

Thus, the only way for the variable to be of any use is to make it
binary, where one category is treatment within a month, and the other is
one month or more.

<br/>

## Chase

```{r}
unique(df$CS_tumor_size)
summary(df$CS_tumor_size, na.rm = TRUE)
sd(df$CS_tumor_size, na.rm = TRUE)
```

```{r}
ggplot(data = df, aes(x = CS_tumor_size))+
  geom_histogram(fill="blue",color="white")+
  labs(x='Tumor Size', y='Frequency', title = 'Tumor Size Histogram')
```

## Frankie

```{r}
#Numerical variable of nodes examined
summary(df$nodes_examined_num)

#Categorical varailable for nodes examined
table(df$nodes_examined_cat)
```

```{r}
#positive nodes numerical variable
summary(df$positive_nodes_num)

#categorical nodes numerical variable
table(df$positive_nodes_cat)
```

<br/>

## Saul

```{r}
# Checking distributions for variables
ggplot(df) +
  geom_histogram(aes(x = age))
```

<br/>

```{r}
ggplot(df) +
  geom_bar(aes(x = race))
```

The vast majority of individuals with a tumor in their testis are white.
We checked the overall demographics of the seer data and found it was
primarily white. Because we do not have access to the counties in which
the individuals reported from, it is difficult to gauge whether this is
an issue or not.

<br/>

```{r}
ggplot(df) +
  geom_histogram(aes(x = survival.months))
```

<br/>

```{r}
ggplot(df, aes(x = marital.status.at.diagnosis, fill = marital.status.at.diagnosis)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        legend.text = element_text(size = 7))
```

Most patients were either married, or single (never married)

<br/>

```{r}
ggplot(df) +
  geom_bar(aes(x = situ.malignant.tumors))
```

Changed to binary (One tumor or More than one tumor)

```{r}
ggplot(df, aes(x = tm.3.1998.2003, fill = tm.3.1998.2003)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        legend.text = element_text(size = 7))
```

Tons of 'NA' values in tumor marker variables, probably in part do them
not spanning all years.

# Modeling

### Survival Analysis

```{r}
df$testi_death <- ifelse(df$death.site == "Testis", 1, 0)
df$SurvObj <- with(df, Surv(survival.months, testi_death == 1))
```

```{r}
km <- survfit(SurvObj ~ 1, data = df, conf.type = "log-log")

km.trt <- survfit(SurvObj ~ Surg.Rad.Seq, data = df, conf.type = "log-log")

ggs <- ggsurvplot(
  fit = survfit(Surv(survival.months, testi_death) ~ Surg.Rad.Seq, data = df), 
  xlab = "Months", 
  ylab = "Overall survival probability") 

ggs$plot <- ggs$plot + ylim(c(0.9, 1))

ggs
```

```{r}
ggs <- ggsurvplot(
  fit = survfit(Surv(survival.months, testi_death) ~ marital.status.at.diagnosis, data = df), 
  xlab = "Months", 
  ylab = "Overall survival probability") 

ggs$plot <- ggs$plot + ylim(c(0.9, 1))

ggs
```

```{r}
mod <- coxph(Surv(survival.months, testi_death) ~ race + Surg.Rad.Seq + Months.diag.to.treat + marital.status.at.diagnosis, data = df)
# %>% gtsummary::tbl_regression(exp = TRUE)
```
The p-vales for all three overall tests (Likelihood, Wald, and Score) are less than .01, indicating the model is significant and rejecting the null hypothesis that all betas are 0. The coefficients of Black and Pacific Islander or Asian are both positive which indicates a higher hazard ratio, but both groups have p-values of 0.140318 > .05 and 0.806804 > .05 respectively signaling they are not statistically significant. The unknown race has a negative coefficient which indicates a lower hazard ratio and is statistically significant with a p-value = 0.000360 < .05. The coefficient of White is negative, which suggests a lower hazard ratio, but is not statistically significant p-value = 0.575 > .05. 

### Checking Cox Proportional Hazards assumptions

```{r}
test.ph <- cox.zph(mod)
```

From the output above, the test is not statistically significant for the
global test as well as each of the covariates. Therefore, we can assume
the proportional hazards.

```{r}
ggcoxzph(test.ph)
```

No pattern with time (supports the use of proportional hazards)

### Model Performance


```{r}
summary(mod)
```

### Interpretations

```{r}
gtsummary::tbl_regression(mod, exp = TRUE)
```

Based on the above regression table, we can interpret the following
information from our model:

-   *Marital status:*

    -   Patients that were married at the time of diagnosis had half
        (0.4, 0.64) the risk of death due to testicular cancer compared
        to patients that were divorced at the time of diagnosis. (p \<
        0.001)

    -   Patients separated from their partner at diagnosis had higher
        risk of death to testicular cancer than divorced patients by a
        factor of 1.56 (1.01, 2.39) (p = 0.044).

    -   These observations point towards a trend where patients that are
        more lonely, or have less emotional support are at more risk of
        death to their diagnosed testicular cancer.
        
        
-   *Months from diagnosis to treatment:* The hazard for death was higher for patients not receiving treatment for at least a month after being diagnosed by a factor of 2.16 (95% CI: 1.90 - 2.45, p= 0.001). The control for this variable were patients treated within a month of the diagnosis.
