---
title: "Jhet: Cleaning"
author: "Jhet Cabigas"
date: "2024-02-08"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(data.table)
library(dplyr)
#read in data
seer <- fread("C:/Users/cjhet/Documents/export.txt", header=TRUE)
```

### Github Repo: <https://github.com/Smoorad99/485-TT>

### Research Questions:

The questions we aim to solve with our data are the following:

-   How does race affect survival rate/months survived for testi cancer?

    -   Exploring the role race plays in testicular cancer survival
        rates could help us address disparities between different races.
        If we find that survival rate of testicular cancer is impacted
        by race, it would push us to explore why this may be. For
        example, it may indicate inequality in healthcare received by
        different races. 

-   How does socioeconomic standing affect survival rate/months survived
    for testi cancer?

    -   Socioeconomic status can impact the quality of healthcare an
        individual has (among other things), which we expect to impact
        testicular cancer survival rate. Exploring this could help us
        address if there is a significant relationship between
        socioeconomic status and testicular cancer survival rate.
        Suppose the difference is significant and the magnitude is
        great. In that case, it may suggest that a policy change could
        be beneficial (i.e., allocate more taxpayer money to healthcare
        for individuals in lower socioeconomic brackets).

-   How does survival rate/months change based on treatment options for
    testi cancer?

    -   Exploring the relationship between survival rate/months and
        treatment method helps us understand which treatment methods are
        most effective. Investigating the quality of life patients
        experience while undergoing different treatments may also help
        us better understand the effectiveness of each treatment.

-   Has the survival rate of testi cancer increased/decreased over
    time? 

    -   Knowing the direction in which the survival rate moves allows us
        to question why it moves in said direction. If the survival rate
        has increased over time, we may ask: Has better
        medicine/treatment options led to this increase in survival
        rate?

## Jhet: Data retrieval, Cols 1-11

### Retrieval:

Using SEERStat, we retrieved a text file containing x observations for
the following variables:

-   race

-   sex

-   year of diagnosis

-   primary site

-   RX Summ (all 6)

-   Reason no cancer-directed surgery

-   Radiation recode

-   Chemotherapy recode

-   Scope of reg lymph nd surg

-   Surgery of other reg/dis sites

-   Site specific surgery

-   Radiation to brain or cns recode?

-   Months from diagnosis to treatment

-   AFP Post-orchiectomy lab value recode?

-   hCG post-orchiectomy range recode

-   LDH post-orchiectomy range recode

-   Number of Examined Pelvic Nodes 

-   Number of Positive Para-Aortic Nodes

-   Number of Positive Pelvic Nodes

-   Peritoneal Cytology

-   Lymph-vascular Invasion

-   CS tumor size (2004-2015) 

-   Regional nodes examined

-   Regional nodes positive

-   CS lymph nodes

-   CS mets at dx

-   CS Tumor Size/Ext Eval

-   CS Reg Node Eval

-   Tumor marker 1

-   Tumor marker 2

-   Tumor marker 3

-   Survival months

-   Survival months flags

-   First malignant primary indicator

-   Total number of in situ/malignant tumors for patient

-   Total number of benign/borderline tumors for patient

-   Race/ethnicity

-   Race and origin recode (NHW, NHB, NHAIAN, NHAPI, Hispanic

-   Age recode with \<1 year olds and 90+

-   Age recode with single ages and 85+

-   Age recode with single ages and 90+

-   Race recode (W, B, AI, API)

-   Marital status at diagnosis

### Data cleaning:

The variables are relatively clean, and only require some recoding as
factor or numeric variables instead of strings. Beyond that, NA's are
recorded as "Blank(s)", which needed to be changed.

```{r}
subset <- seer %>% select(1:11)

subset[subset == "Blank(s)"] <- NA
subset <- subset %>% mutate(
  Sex = as.factor(Sex),
  `Race and origin recode (NHW, NHB, NHAIAN, NHAPI, Hispanic)` = as.factor(`Race and origin recode (NHW, NHB, NHAIAN, NHAPI, Hispanic)`),
  `Age recode with <1 year olds` = as.factor(`Age recode with <1 year olds`), # REMOVE
  `RX Summ--Scope Reg LN Sur (2003+)` = as.factor(`RX Summ--Scope Reg LN Sur (2003+)`),
  `RX Summ--Surg Oth Reg/Dis (2003+)` = as.factor(`RX Summ--Surg Oth Reg/Dis (2003+)`),
  `RX Summ--Surg/Rad Seq` = as.factor(`RX Summ--Surg/Rad Seq`),
  `RX Summ--Reg LN Examined (1998-2002)` = as.numeric(`RX Summ--Reg LN Examined (1998-2002)`),
  `RX Summ--Systemic/Sur Seq (2007+)` = as.factor(`RX Summ--Systemic/Sur Seq (2007+)`),
  `Months from diagnosis to treatment` = as.numeric(`Months from diagnosis to treatment`)
)
```

## Chase: Cols 12-22

Used tables and frequency tables to analyze the columns. One variable
was factored, every entry is listed as "Blank(s)" in
`Radiation to Brain or CNS Recode (1988-1997)` and
`Site specific surgery (1973-1997 varying detail by year and site)`.

```{r}
names(seer)
```

```{r}
table(seer$`AFP Post-Orchiectomy Lab Value Recode (2010+)`)
factor(seer$`hCG Post-Orchiectomy Range Recode (2010+)`,labels=c("5,000-50,000", "Above normal and less than 5,000 MIU/mL", "Blank(s)", "Greater than 50,000" , "Blank(s)", "Blank(s)", "Blank(s)", "normal"))
```

```{r}
table(seer$`AFP Post-Orchiectomy Lab Value Recode (2010+)`)

```

```{r}
table(seer$`Surgery of oth reg/dis sites (1998-2002)`)
table(seer$`Scope of reg lymph nd surg (1998-2002)`)
```

```{r}
table(seer$`Chemotherapy recode (yes, no/unk)`)
table(seer$`Radiation recode`)
table(seer$`Reason no cancer-directed surgery`)
```

```{r}
unique(seer$`Primary Site`)
```

## Frankie: Cols 23-32

Going through variables 23-32 from our original data set to verify
quality in data. Making sure to remove columns that have too many
missing variables. Removed columns 23-26 becasue they contained too many
blanks.

## Saul: Cols 33-47

I checked variables for suspicious values, changed variables from
strings to numeric (when appropriate), and renamed variables for ease of
use. I then dropped a few variables that did not add any useful
information.

```{r}
a <- seer[, 33:47] # Subsetting data
b <- a # For reference
rm(seer)

colnames(a) <- gsub(" ", ".", colnames(a)) # replace spaces with periods
colnames(b) <- gsub(" ", ".", colnames(b))

# Three age variables... which do we want
table(a$`Age.recode.with.<1.year.olds.and.90+`) # looks clean

table(a$`Age.recode.with.single.ages.and.85+`) # if we decide we want this remove " years" and convert to numeric

table(a$`Age.recode.with.single.ages.and.90+`) # same as above

a$`Age.recode.with.single.ages.and.90+` <- gsub(" years", "",a$`Age.recode.with.single.ages.and.90+`) 

table(a$`Age.recode.with.single.ages.and.90+`)

a <- a %>% 
  mutate(
    `CS.Reg.Node.Eval.(2004-2015)` = as.numeric(`CS.Reg.Node.Eval.(2004-2015)`),
    `CS.Tumor.Size/Ext.Eval.(2004-2015)` = as.numeric(`CS.Tumor.Size/Ext.Eval.(2004-2015)`)
    )


table(a$`CS.Reg.Node.Eval.(2004-2015)`)
table(b$`CS.Reg.Node.Eval.(2004-2015)`) # Check recode

table(a$`CS.Tumor.Size/Ext.Eval.(2004-2015)`)
table(b$`CS.Tumor.Size/Ext.Eval.(2004-2015)`) # Check recode


table(a$Marital.status.at.diagnosis) # looks clean... Were non-answers replaced with 'Unknown'?
table(a$`Race.recode.(W,.B,.AI,.API)`) # Looks clean

table(a$`Race/ethnicity`) #Looks clean

a$Survival.months <- as.numeric(a$Survival.months)

table(a$Survival.months)
table(b$Survival.months) # Check recode

table(a$Survival.months.flag) # Looks clean

table(a$`Total.number.of.benign/borderline.tumors.for.patient`) # Looks clean, not sure if this variableis worth keeping due the distribution of responses

table(a$`Total.number.of.in.situ/malignant.tumors.for.patient`)

a$`Total.number.of.in.situ/malignant.tumors.for.patient` <- as.numeric(a$`Total.number.of.in.situ/malignant.tumors.for.patient`)

table(a$`Total.number.of.in.situ/malignant.tumors.for.patient`)
table(b$`Total.number.of.in.situ/malignant.tumors.for.patient`) # Checking recode

table(a$`Tumor.marker.1.(1990-2003)`) # Not sure how to treat these tumor marker variables, need a better understanding of what they are before deciding

table(a$`Tumor.marker.2.(1990-2003)`)

table(a$`Tumor.marker.3.(1998-2003)`)


## Changing names of variables we want to keep
a <- a %>% 
  mutate(
    age = `Age.recode.with.single.ages.and.90+`,
    cs.reg.node.eval.04.15 = `CS.Reg.Node.Eval.(2004-2015)`,
    cs.tumor.size.ext.eval.04.15 = `CS.Tumor.Size/Ext.Eval.(2004-2015)`,
    race.recode = `Race.recode.(W,.B,.AI,.API)`,
    race.ethnicity = `Race/ethnicity`,
    survival.months = Survival.months,
    survival.months.flag = Survival.months.flag,
    marital.status.at.diagnosis = Marital.status.at.diagnosis,
    benign.borderline.tumors = `Total.number.of.benign/borderline.tumors.for.patient`,
    situ.malignant.tumors = `Total.number.of.in.situ/malignant.tumors.for.patient`,
    tm.1.1990.2003 = `Tumor.marker.1.(1990-2003)`,
    tm.2.1990.2003 = `Tumor.marker.2.(1990-2003)`,
    tm.3.1998.2003 = `Tumor.marker.3.(1998-2003)`)

saul_semiclean_vars <- a %>% select(age, cs.reg.node.eval.04.15, cs.tumor.size.ext.eval.04.15,
                          race.recode, race.ethnicity, survival.months,
                          survival.months.flag, Marital.status.at.diagnosis, 
                          benign.borderline.tumors, situ.malignant.tumors,
                          tm.1.1990.2003, tm.2.1990.2003, tm.3.1998.2003)
```
